Cost- and Deadline-constrained Scheduling of Scientific Workflow Ensembles in IaaS Clouds

### 2012-01-23 paper/paper.tex
change ref to line in alg 2
updated data in table 2
added last paragraph

\section{Related Work}
\label{sec:related}
General policy and rule-based approaches to dynamic provisioning (e.g. Amazon
Auto Scaling\footnote{\url{http://aws.amazon.com/autoscaling}} and
RightScale\footnote{\url{http://www.rightscale.com}}) allow adjusting the size
of resource pool based on metrics related to infrastructure and application.
A typical infrastructure-specific metric is system load, whereas
application-specific metrics include response time and length of task or
request queue. It is possible to set thresholds and limits to tune the behavior
of these autoscaling systems, but no support for complex applications is provided.

@@

Our work is related to the strategies for deadline-constrained cost-minimization
workflow scheduling, developed for utility grid systems. However, our problem is
different from \cite{Yu2005, Abrishami2010} in that we consider ensembles of workflows in
IaaS clouds, which allow one to provision resources on a per-hour billing model,
rather than utility grids, which allow one to choose from a pool of existing
resources with a per-job billing model. Our work is also different from
cloud-targeted autoscaling solution~\cite{Mao2011} in that we consider ensembles
of workflows rather than unpredictable workloads containing workflows. We also consider budget constraints
rather than cost minimization as a goal. In other words, we assume that there is
more work to be done than the available budget, so some work must be rejected.
Therefore cost is not something we optimize (i.e. an objective), but rather a constraint.

@@

\begin{algorithm}
\caption{Priority-based scheduling algorithm for SPDS}
\label{alg:ds}
\begin{algorithmic}[1]
\Procedure{Schedule}{}
    \State $P\gets$ empty priority queue
	\State $IdleVMs\gets$ set of idle VMs
	\For{root task $t$ in all workflows}
    	\State \Call{Insert}{$t,P$} 
    \EndFor
    \While{deadline not reached}
    	\While{$IdleVMs \neq \emptyset\ and\ P \neq \emptyset$}
    		\State $v\gets$ \Call{SelectRandom}{$IdleVMs$}
    		\State $t\gets$ \Call{Pop}{$P$}
    		\State \Call{Submit}{$t,v$}
    	\EndWhile
    	\State Wait for task $t$ to finish on VM $v$
    	\State Update $P$ with ready children of $t$
		\State \Call{Insert}{$v,IdleVMs$}
    \EndWhile
\EndProcedure
\end{algorithmic} 
\end{algorithm}

@@

\begin{algorithm}
\caption{Dynamic provisioning algorithm for DPDS}
\label{alg:prov}
\begin{algorithmic}[1]
\Require $c$: consumed budget; $b$: total budget; $d$: deadline; $p$: price;
$t$: current time; $u_h$: upper utilization threshold; $u_l$: lower utilization
threshold; $v_{max}$: maximum number of VMs
\Procedure{Provision}{}
	\State $V_R\gets$ set of running VMs
    \State $V_C\gets$ set of VMs completing billing cycle
    \If{$ b - c < |V_C| * p\ or\ t > d $ }
    	\State $n_T\gets |V_R| - \lfloor(b-c)/p\rfloor$
    	\State $V_T\gets$ select $n_T$ VMs to terminate from $V_C$
    	\State \Call{Terminate}{$V_T$} \label{l:terminate1}
    \Else 
		\State $u\gets$ current VM utilization
    	\If{$u>u_h$ and $|V_R| < v_{max}*N_{VM}$}
    		\State \Call{Start}{$new\ VM$}
    	\ElsIf{$u<u_l$}
    		\State $V_I\gets$ set of idle VMs
    		\State $n_T\gets \lceil|V_I|/2\rceil$ \label{l:nT2}
			\State $V_T\gets$ select $n_T$ VMs to terminate from $V_I$
    		\State \Call{Terminate}{$V_T$} \label{l:terminate2}
    	\EndIf 
    \EndIf
\EndProcedure
\end{algorithmic} 
\end{algorithm}

The set of VMs completing their billing cycle is determined by considering both the
provisioner interval, and the termination delay of the provider. This guarantees 
that VMs can be terminated before they start the next billing cycle and prevents the
budget from being overrun. The VMs terminated in line \ref{l:terminate1} of Algorithm~\ref{alg:prov} are the ones 
that would overrun the budget if not terminated in the current provisioning cycle. 
The VMs terminated on line \ref{l:nT2} are chosen to increase the resource utilization
to the desired threshold. In order to prevent instances from being
terminated too quickly, potentially wasting resources that have already been paid for
but could be used later, no more than half of the idle resources are terminated during
each provisioning cycle. To avoid an uncontrolled increase in the number of
instances, which may happen in the case of highly parallel workflows, the provisioner 
will not start a new VM if the number of running VMs is greater than the product of
$N_{VM}$ and an autoscaling factor, $v_{max}$ that has to be set manually.

@@

In order to address this issue, the Workflow-Aware DPDS (WA-DPDS) algorithm extends 
DPDS by introducing a workflow admission procedure. The admission procedure is
invoked whenever WA-DPDS sees the first task of a new workflow at the head of the 
priority queue (i.e. when no other tasks from the workflow have been scheduled yet). 
The admission procedure---shown in Algorithm~\ref{alg:wa-dpds}---estimates whether there is 
enough budget remaining to admit the new workflow; if there is not, then the 
workflow is rejected and its tasks are removed from the queue. This algorithm 
compares the current cost (consumed budget) and remaining budget, taking
into account the cost of currently running VMs, and the cost of workflows 
that have already been admitted. It adds a small safety margin to  avoid going over
the budget.

\begin{algorithm}
\caption{Workflow admission algorithm for WA-DPDS}
\label{alg:wa-dpds}
\begin{algorithmic}[1]
\Require $w$: workflow; $b$: budget; $c$: current cost
\Procedure{Admit}{$w,b,c$}
    \State $r_n\gets b-c$ \Comment{Budget remaining for new VMs}
    \State $r_c\gets $cost committed to VMs that are running
    \State $r_a\gets $cost to complete workflows previously admitted
	\State $r_m\gets 0.1$ \Comment{Safety margin}
	\State $r_b\gets r_n+r_c-r_a-r_m$ \Comment{Budget remaining}
	\State $c_w\gets$ \Call{EstimateCost}{w}
    \If{$c_w<r_b$}
    	\State \textbf{return} $TRUE$
    \Else
    	\State \textbf{return} $FALSE$
	\EndIf    	 
\EndProcedure
\end{algorithmic} 
\end{algorithm}

@@

\begin{algorithm}
\caption{Ensemble planning algorithm for SPSS}
\label{alg:admit}
\begin{algorithmic}[1]
\Require $W$: workflow ensemble; $b$: budget; $d$: deadline
\Ensure Schedule as much of $W$ as possible given $b$ and $d$
\Procedure{PlanEnsemble}{$W,b,d$}
    \State $P\gets \emptyset$ \Comment{Current plan}
    \State $A\gets \emptyset$ \Comment{Set of admitted DAGs}
    \For{$w\ in\ W$}
        \State $P^\prime \gets$\ \Call{PlanWorkflow}{$w,P,d$}
        \If{$Cost(P^\prime) \le b$}
            \State $P\gets\ P^\prime$ \Comment{Accept new plan}
            \State $A \gets A\ +\ w$ \Comment{Admit w}
        \EndIf
    \EndFor
    \State \textbf{return} $P,A$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

@@

\begin{algorithm}
\caption{Workflow planning algorithm for SPSS}
\label{alg:planworkflow}
\begin{algorithmic}[1]
\Require $w$: workflow; $P$: current plan; $d$: deadline
\Ensure Create plan for $w$ that minimizes cost and meets deadline $d$
\Procedure{PlanWorkflow}{$w,P,d$}
    \State $P^\prime\gets$ copy of $P$
    \State \Call{DeadlineDistribution}{w,d}
    \For{$t\ in\ w\ sorted\ by\ DL(t)$}
        \State $v \gets$ VM that minimizes cost and start time of t
        \If{$FinishTime(t,v) < DL(t)$}
            \State Schedule(t,v)
        \Else
            \State Provision a new VM v
            \State Schedule(t,v)
        \EndIf
    \EndFor
    \State \textbf{return} $P^\prime$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

\begin{figure}[htb] 
\centering
\includegraphics[width=1.0\columnwidth]{figures/spss-gantt}
 \caption{Example of schedule generated using SPSS algorithm. Tasks are colored by workflow. }
\label{fig:spss-example}
\end{figure}

@@

Figure~\ref{fig:number-complete-pareto} shows the number of the workflows
completed for two budget values for each application ensemble. The budget values
selected correspond to two extreme cases: {\em small budget} (top) and {\em
large budget} (bottom). The small budgets permit the algorithms to complete 
only a small number of workflows from the ensemble, while the larger budgets 
enable the algorithms to finish almost all the workflows. The rangesIn three of
five cases the SPSS algorithm performs better for tighter constraints. When the 
budget is low, or the deadline is short, the static procedure is able to more 
efficiently pack workflow tasks onto limited resources. This is the case for 
the LIGO, Epigenomics and SIPHT ensembles for small budgets, and for large
budgets with short deadlines. The exceptions are the Montage and CyberShake 
ensembles,  which consist of many fine-grained tasks. In these applications, 
as well for the other ensembles with large budgets and deadlines, the simple 
dynamic scheduling techniques perform equally well or better. The reason for 
this behavior is that the large number of tasks guarantees that resources 
are almost never idle regardless of how the tasks are scheduled. We also 
observe that WA-DPDS performs better than the workflow-unaware version, DPDS. 
This means that for the workflows we tested the simple admission procedure based 
on estimation of workflow cost does not degrade the solution, but rather in 
many cases it enables larger workflows that would lead to budget overrun to
be rejected. Thus it can save the space for smaller workflows to complete.

@@

(a) Number of workflows completed $N_c$:
\medskip
\begin{tabular}{r|cccccc}
 & M & C & L & E & S & All\tabularnewline
\hline
DPDS      &   78 &  93 & 161 &  24  & 30 & 386\tabularnewline
WA-DPDS   &    78  & 97 & 179  & 58  & 28 & 440\tabularnewline
SPSS     &    78  & 79 & 195 &  75 &  50 & 477\tabularnewline
\end{tabular}
\medskip

(b) Total amount of computing time completed $T_C$:
\medskip
\begin{tabular}{r|cccccc}
 & M & C & L & E & S & All\tabularnewline
\hline
DPDS      &   78 &  93 & 161 &  24 &  30 & 386\tabularnewline
WA-DPDS   &    78  & 97 & 179 &  46  & 28 & 428\tabularnewline
SPSS     &    78 &  77 & 195  & 75 &  50 & 475\tabularnewline
\end{tabular}
\medskip

(c) Exponential score $s$:
\medskip
\begin{tabular}{r|cccccc}
 & M & C & L & E & S & All\tabularnewline
\hline
DPDS      &   78  & 93 & 161 &  24 &  30 & 386\tabularnewline
WA-DPDS   &    78 &  97 & 179  & 46  & 28 & 428\tabularnewline
SPSS     &    78 &  77 & 195 &  75  & 50 & 475\tabularnewline
\end{tabular}
\medskip

@@

In order to compare the efficiency of our algorithms, we counted the number of times
each algorithm achieved the best result for all simulated applications, deadlines
and budgets. Table~\ref{tab:num-dags-pareto} shows these summary values for 
Pareto-distributed workflows. Note that all algorithms that achieve the best 
result are credited, so the sum for a given application can be greater than the
total number of scenarios. These results depend on the metric: number of
workflows completed does not count the priorities or sizes of workflows, which
means that it happens more often that all the algorithms achieve the same best
result. However, it does not mean that all algorithms complete the same
workflows: that is why the values in (a), (b) and (c) are different. The 
exponential score and total amount of computing time completed give the same
rankings, since in our test ensembles larger workflows were given higher
priorities. It can be observed based on the sampled problem space that the
workflow-aware algorithms achieve the highest score in $\sim$2--3 times more 
cases than the simple DPDS algorithm, which does not take into account workflow 
structure or size. SPSS performs the best in $\sim$20\% more cases than WA-DPDS,
including the Montage and CyberShake ensembles, for which SPSS does not 
perform as well as WA-DPDS.

@@

\section{Conclusion and Future Work}
\label{sec:conclusions}

@@

>>>>

\section{Related Work}
\label{sec:related}
General policy and rule-based approaches to dynamic provisioning (e.g. Amazon
Auto Scaling\footnote{\url{http://aws.amazon.com/autoscaling}} and
RightScale\footnote{\url{http://www.rightscale.com}}) allow adjusting the size
of resource pool based on metrics related to infrastructure and application.
A typical infrastruc\-ture-specific metric is system load, whereas
application-speci\-fic metrics include response time and length of task or
request queue. It is possible to set thresholds and limits to tune the behavior
of these autoscaling systems, but no support for complex applications is provided.

@@

Our work is related to the strategies for deadline-con\-strained
cost-minimization workflow scheduling, developed for utility grid systems. However, our problem is
different from \cite{Yu2005, Abrishami2010} in that we consider ensembles of workflows in
IaaS clouds, which allow one to provision resources on a per-hour billing model,
rather than utility grids, which allow one to choose from a pool of existing
resources with a per-job billing model. Our work is also different from
cloud-targeted autoscaling solution~\cite{Mao2011} in that we consider ensembles
of workflows rather than unpredictable workloads containing workflows. We also consider budget constraints
rather than cost minimization as a goal. In other words, we assume that there is
more work to be done than the available budget, so some work must be rejected.
Therefore cost is not something we optimize (i.e. an objective), but rather a constraint.

@@

\begin{algorithm}[tb]
\caption{Priority-based scheduling algorithm for SPDS}
\label{alg:ds}
\begin{algorithmic}[1]
\Procedure{Schedule}{}
    \State $P\gets$ empty priority queue
	\State $IdleVMs\gets$ set of idle VMs
	\For{root task $t$ in all workflows}
    	\State \Call{Insert}{$t,P$} 
    \EndFor
    \While{deadline not reached}
    	\While{$IdleVMs \neq \emptyset\ and\ P \neq \emptyset$}
    		\State $v\gets$ \Call{SelectRandom}{$IdleVMs$}
    		\State $t\gets$ \Call{Pop}{$P$}
    		\State \Call{Submit}{$t,v$}
    	\EndWhile
    	\State Wait for task $t$ to finish on VM $v$
    	\State Update $P$ with ready children of $t$
		\State \Call{Insert}{$v,IdleVMs$}
    \EndWhile
\EndProcedure
\end{algorithmic} 
\end{algorithm}

@@

\begin{algorithm}[tb]
\caption{Dynamic provisioning algorithm for DPDS}
\label{alg:prov}
\begin{algorithmic}[1]
\Require $c$: consumed budget; $b$: total budget; $d$: deadline; $p$: price;
$t$: current time; $u_h$: upper utilization threshold; $u_l$: lower utilization
threshold; $v_{max}$: maximum number of VMs
\Procedure{Provision}{}
	\State $V_R\gets$ set of running VMs
    \State $V_C\gets$ set of VMs completing billing cycle
    \If{$ b - c < |V_C| * p\ or\ t > d $ }
    	\State $n_T\gets |V_R| - \lfloor(b-c)/p\rfloor$
    	\State $V_T\gets$ select $n_T$ VMs to terminate from $V_C$
    	\State \Call{Terminate}{$V_T$} \label{l:terminate1}
    \Else 
		\State $u\gets$ current VM utilization
    	\If{$u>u_h$ and $|V_R| < v_{max}*N_{VM}$}
    		\State \Call{Start}{$new\ VM$}
    	\ElsIf{$u<u_l$}
    		\State $V_I\gets$ set of idle VMs
    		\State $n_T\gets \lceil|V_I|/2\rceil$ \label{l:nT2}
			\State $V_T\gets$ select $n_T$ VMs to terminate from $V_I$
    		\State \Call{Terminate}{$V_T$} \label{l:terminate2}
    	\EndIf 
    \EndIf
\EndProcedure
\end{algorithmic} 
\end{algorithm}

The set of VMs completing their billing cycle is determined by considering both the
provisioner interval, and the termination delay of the provider. This guarantees 
that VMs can be terminated before they start the next billing cycle and prevents the
budget from being overrun. The VMs terminated in line \ref{l:terminate1} of Algorithm~\ref{alg:prov} are the ones 
that would overrun the budget if not terminated in the current provisioning cycle. 
The VMs terminated in line \ref{l:terminate2} are chosen to increase the resource
utilization to the desired threshold. In order to prevent instances from being
terminated too quickly, potentially wasting resources that have already been paid for
but could be used later, no more than half of the idle resources are terminated during
each provisioning cycle. To avoid an uncontrolled increase in the number of
instances, which may happen in the case of highly parallel workflows, the provisioner 
will not start a new VM if the number of running VMs is greater than the product of
$N_{VM}$ and an autoscaling factor, $v_{max}$ that has to be set manually.

@@

In order to address this issue, the Workflow-Aware DPDS (WA-DPDS) algorithm extends 
DPDS by introducing a workflow admission procedure. The admission procedure is
invoked whenever WA-DPDS sees the first task of a new workflow at the head of the 
priority queue (i.e. when no other tasks from the workflow have been scheduled yet). 
The admission procedure -- shown in Algorithm~\ref{alg:wa-dpds} -- estimates
whether there is enough budget remaining to admit the new workflow; if there is not, then the 
workflow is rejected and its tasks are removed from the queue. This algorithm 
compares the current cost (consumed budget) and remaining budget, taking
into account the cost of currently running VMs, and the cost of workflows 
that have already been admitted. It adds a small safety margin to  avoid going over
the budget.

\begin{algorithm}[tb]
\caption{Workflow admission algorithm for WA-DPDS}
\label{alg:wa-dpds}
\begin{algorithmic}[1]
\Require $w$: workflow; $b$: budget; $c$: current cost
\Procedure{Admit}{$w,b,c$}
    \State $r_n\gets b-c$ \Comment{Budget remaining for new VMs}
    \State $r_c\gets $cost committed to VMs that are running
    \State $r_a\gets $cost to complete workflows previously admitted
	\State $r_m\gets 0.1$ \Comment{Safety margin}
	\State $r_b\gets r_n+r_c-r_a-r_m$ \Comment{Budget remaining}
	\State $c_w\gets$ \Call{EstimateCost}{w}
    \If{$c_w<r_b$}
    	\State \textbf{return} $TRUE$
    \Else
    	\State \textbf{return} $FALSE$
	\EndIf    	 
\EndProcedure
\end{algorithmic} 
\end{algorithm}

@@

\begin{algorithm}[tb]
\caption{Ensemble planning algorithm for SPSS}
\label{alg:admit}
\begin{algorithmic}[1]
\Require $W$: workflow ensemble; $b$: budget; $d$: deadline
\Ensure Schedule as much of $W$ as possible given $b$ and $d$
\Procedure{PlanEnsemble}{$W,b,d$}
    \State $P\gets \emptyset$ \Comment{Current plan}
    \State $A\gets \emptyset$ \Comment{Set of admitted DAGs}
    \For{$w\ in\ W$}
        \State $P^\prime \gets$\ \Call{PlanWorkflow}{$w,P,d$}
        \If{$Cost(P^\prime) \le b$}
            \State $P\gets\ P^\prime$ \Comment{Accept new plan}
            \State $A \gets A\ +\ w$ \Comment{Admit w}
        \EndIf
    \EndFor
    \State \textbf{return} $P,A$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}[tb]
\caption{Workflow planning algorithm for SPSS}
\label{alg:planworkflow}
\begin{algorithmic}[1]
\Require $w$: workflow; $P$: current plan; $d$: deadline
\Ensure Create plan for $w$ that minimizes cost and meets deadline $d$
\Procedure{PlanWorkflow}{$w,P,d$}
    \State $P^\prime\gets$ copy of $P$
    \State \Call{DeadlineDistribution}{w,d}
    \For{$t\ in\ w\ sorted\ by\ DL(t)$}
        \State $v \gets$ VM that minimizes cost and start time of t
        \If{$FinishTime(t,v) < DL(t)$}
            \State Schedule(t,v)
        \Else
            \State Provision a new VM v
            \State Schedule(t,v)
        \EndIf
    \EndFor
    \State \textbf{return} $P^\prime$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

@@

\begin{figure}[tb] 
\centering
\includegraphics[width=1.0\columnwidth]{figures/spss-gantt}
 \caption{Example of schedule generated using SPSS algorithm. Tasks are colored by workflow. }
\label{fig:spss-example}
\end{figure}

@@

Figure~\ref{fig:number-complete-pareto} shows the number of the workflows
completed for two budget values for each application ensemble. The budget values
selected correspond to two extreme cases: {\em small budget} (top) and {\em
large budget} (bottom). The small budgets permit the algorithms to complete 
only a small number of workflows from the ensemble, while the larger budgets 
enable the algorithms to finish almost all the workflows. In three of
five cases the SPSS algorithm performs better for tighter constraints. When the 
budget is low, or the deadline is short, the static procedure is able to more 
efficiently pack workflow tasks onto limited resources. This is the case for 
the LIGO, Epigenomics and SIPHT ensembles for small budgets, and for large
budgets with short deadlines. The exceptions are the Montage and CyberShake 
ensembles,  which consist of many fine-grained tasks. In these applications, 
as well for the other ensembles with large budgets and deadlines, the simple 
dynamic scheduling techniques perform equally well or better. The reason for 
this behavior is that the large number of tasks guarantees that resources 
are almost never idle regardless of how the tasks are scheduled. We also 
observe that WA-DPDS performs better than the workflow-unaware version, DPDS. 
This means that for the workflows we tested the simple admission procedure based 
on estimation of workflow cost does not degrade the solution, but rather in 
many cases it enables larger workflows that would lead to budget overrun to
be rejected. Thus it can save the space for smaller workflows to complete.

@@

(a) Number of workflows completed $N_c$:
\medskip
\begin{tabular}{r|cccccc}
 & M & C & L & E & S & All\tabularnewline
\hline
DPDS      &   79  & 92  & 170 &  24 &  29 & 394\tabularnewline
WA-DPDS   &   82  & 100 & 180 &  58 &  27 & 447\tabularnewline
SPSS      &   72  & 77  & 195 &  75 &  50 & 469 \tabularnewline
\end{tabular}
\medskip

(b) Total amount of computing time completed $T_C$:
\medskip
\begin{tabular}{r|cccccc}
 & M & C & L & E & S & All\tabularnewline
\hline
DPDS      &   79  & 92  & 170 &  24 &  29 & 394\tabularnewline
WA-DPDS   &   82  & 100 & 180 &  46 &  27 & 435\tabularnewline
SPSS      &   72  & 75  & 195 &  75 &  50 & 467 \tabularnewline
\end{tabular}
\medskip

(c) Exponential score $s$:
\medskip
\begin{tabular}{r|cccccc}
 & M & C & L & E & S & All\tabularnewline
\hline
DPDS      &   79  & 92  & 170 &  24 &  29 & 394\tabularnewline
WA-DPDS   &   82  & 100 & 180 &  46 &  27 & 435\tabularnewline
SPSS      &   72  & 75  & 195 &  75 &  50 & 467 \tabularnewline
\end{tabular}
\medskip

@@

In order to compare the efficiency of our algorithms, we counted the number of times
each algorithm achieved the best result for all simulated applications, deadlines
and budgets. Table~\ref{tab:num-dags-pareto} shows these summary values for 
Pareto-distri\-buted workflows. Note that all algorithms that achieve the best 
result are credited, so the sum for a given application can be greater than the
total number of scenarios. These results depend on the metric: number of
workflows completed does not count the priorities or sizes of workflows, which
means that it happens more often that all the algorithms achieve the same best
result. However, it does not mean that all algorithms complete the same
workflows: that is why the values in (a), (b) and (c) are different. The 
exponential score and total amount of computing time completed give the same
rankings, since in our test ensembles larger workflows were given higher
priorities. It can be observed based on the sampled problem space that the
workflow-aware algorithms achieve the highest score in $\sim$2--3 times more 
cases than the simple DPDS algorithm, which does not take into account workflow 
structure or size. SPSS performs the best in $\sim$20\% more cases than WA-DPDS,
including the Montage and CyberShake ensembles, for which SPSS does not 
perform as well as WA-DPDS.

@@

\section{Conclusions and Future Work}
\label{sec:conclusions}

@@

We believe that the results of our study can be applied in practice to develop 
tools that assist researchers in planning their large-scale computational
experiments. The estimates of costs, runtimes and number of workflows completed
that can be obtained from both the static algorithms and from the simulation
runs, constitute valuable hints for the planning decisions and assessing the
associated trade-offs. For this purpose our future work includes also further
extension of our experimental studies~\cite{Juve2009,vockler11} of workflows on
available cloud infrastructures to develop more accurate performance and cost
models of the applications. This should lead to better understanding of the
problems and costs of doing science in the cloud.
